D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-01-25 ootavacuous.tex
January 25th reorganization, part 1
Based on Alan Stern's feedback up through moving from (sdep U rf) to
(sdep U rfe).

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This section provides a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick review of a few of the challenges for general definitions of
``OOTA cycle''.

@@

OOTA occurs when a group of threads load from each others' stores,
but where each thread's store depends on the value returned by that
thread's load.
A given out-of-thin-air value passes around the resulting cycle.
Because there is a cycle, on a real system, at least one of the links
must be atemporal, that is, either a load returns a value before
that value is stored, or a store's address or value is computed before
a prior load returns a value used in that computation.

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle.
The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access, in this case, \co{x}
and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

Other OOTA-related litmus tests may be found in
Appendix~\ref{app:OOTA-Related Litmus Tests}.

@@

This paper therefore drops the question of how OOTA can be avoided and
instead focuses on proving that OOTA cannot occur.

@@

The C++ standard says this in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

@@

Others have defined an OOTA cycle as a cycle in sdep $\cup$ rf, where
sdep is the set of semantic dependencies within each thread and rf is
the set of store-to-load links, whether intra-thread (rfi) or between
threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}
This is a fine definition, and is consistent with the words in the C++
standard, except that it simply shunts all the complexity onto the
term ``semantic dependency''.

We deviate slightly from this tradition by instead defining an OOTA cycle
as a cycle in sdep $\cup$ rfe (or \co{acyclic(sdep ; rfe)} in \co{herd7}
notation), where rfe is the set of store-to-load links, but excluding
intra-thread rfi links, which are instead considered to be part of sdep.
Although this does shunt even more complexity onto the term
``semantic dependency'', it also enables us to bring to bear
laws of physics and properties of hardware onto correct real-world
C++ implementations.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}
Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.

@@

@@

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
revises
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4}
so that the value stored to \co{y} is forced to be at most the
value 17, and the value stored to \co{z} is forced to be at least the
value 17.
This of course means that the store to \co{z} must always be the value
17, as can be seen from the \co{herd7} output.
However, a single-thread-at-a-time analysis could not prove this.

>>>>

@@@ This section provides a brief overview of the OOTA problem, followed
by an equally brief summary of prior work in this area, and ending
with a quick review of a few of the challenges for general definitions of
``OOTA cycle''.

@@

OOTA occurs when a group of threads load from each others' stores to
each of a group of objects, but where each thread's store depends on
the value returned by that thread's load.
A given out-of-thin-air value passes around the resulting \emph{OOTA cycle}.

\subsubsection{Simple Reordering}
\label{sec:Simple Reordering}

However, it is important to distinguish OOTA cycles from the simple
reordering exemplified by
Listing~\ref{lst:Simple Reordering}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}.
Both the C++ implementation\footnote{
	This paper will use ``C++ implementation'' both excluding the
	CPU, as here, and including the CPU.
	Explicit disambiguation is added when necessary.}
and the CPU are within their rights to reorder lines~12 and~13, which
can result in all of \co{x}, \co{y}, \co{r1}, and \co{r2} having the
value 42 as follows:

\begin{enumerate}
\item   Line~13 stores 42 to \co{x}.
\item   Line~6 loads 42 from \co{x} into \co{r1}.
\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
\item   Line~12 loads 42 from \co{y} to \co{r2}.
\end{enumerate}

Current C++ implementations can and do exhibit this simple-reordering
behavior.

\subsubsection{Simple OOTA Cycle}
\label{sec:Simple OOTA Cycle}

In contrast,
Listing~\ref{lst:Simple OOTA}~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
shows a simple example where an OOTA cycle might result in all of \co{x},
\co{y}, \co{r1}, and \co{r2} have final values of 42, despite the fact
that there is nothing in the code or in the initial values that would
support such an outcome:

\begin{enumerate}
\item   Line~6 loads from \co{x} into \co{r1}, somehow obtaining the
	value 42, never mind the \co{x}'s initial value of zero.
\item   Line~7 stores \co{r1}, and thus 42, to \co{y}.
\item   Line~12 loads 42 from \co{y} to \co{r2}.
\item   Line~13 stores 42 to \co{x}, justifying the value loaded by
	line~6 after the fact.
\end{enumerate}

Fortunately, the C++ standard prohibits such OOTA cycles in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

This paper will follow P2055R0~\cite{PaulEMcKenney2020RelaxedGuideRelaxed}
in using the term \emph{full C++} to denote the standard including the
above prohibition, and the term \emph{strict C++} to denote a hypothetical
standard that excluded this prohibition, but was otherwise identical
to full C++.
Unqualified \emph{C++} means full C++.

The next section looks at how prior work has refined this prohibition.

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   unsigned long r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   unsigned long r2 = y.load(memory_order_relaxed);
13   x.store(42, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple Reordering}
\label{lst:Simple Reordering}
\end{listing}

\begin{listing}[tbp]
\begin{verbatim}
 1 atomic<int> x(0);
 2 atomic<int> y(0);
 3
 4 void thread1()
 5 {
 6   unsigned long r1 = x.load(memory_order_relaxed);
 7   y.store(r1, memory_order_relaxed);
 8 }
 9
10 void thread2()
11 {
12   unsigned long r2 = y.load(memory_order_relaxed);
13   x.store(r2, memory_order_relaxed);
14 }
\end{verbatim}
\caption{Simple OOTA}
\label{lst:Simple OOTA}
\end{listing}

@@

This paper therefore digs more deeply into OOTA cycles in the light
of these real-world constraints.

\subsection{Code-Analysis Tool}
\label{sec:Code-Analysis Tool}

This paper will use the \co{herd7} tool to analyze fragments of C++ code.
This tool carries out the moral equivalent of full state-space searches
of concurrent code fragments, and further has some understanding of OOTA.

Listing~\ref{lst:OOTA Cycle}
shows a code fragment that in strict C++ would have an OOTA cycle.
This section describes this fragment, thus giving an overview of
the \co{herd7} tool.

The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access, in this case, \co{x}
and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

The earlier examples shown in
Listings~\ref{lst:Simple Reordering}
and~\ref{lst:Simple OOTA}
can be recast as similar \co{herd7} litmus tests, which areas shown in
Listing~\ref{lst:Simple Reordering as Litmus Test}
on page~\pageref{lst:Simple Reordering as Litmus Test}
and Listing~\ref{lst:Causality Test Case 4}
on page~\pageref{lst:Causality Test Case 4}, respectively.
Other OOTA-related litmus tests may be found in
Appendix~\ref{app:OOTA-Related Litmus Tests}.

\section{Semantic Dependencies, Executions, and OOTA}
\label{sec:Semantic Dependencies; Executions; and OOTA}

@@@

\subsection{Semantic Dependencies}
\label{sec:Semantic Dependencies}

The standard's prohibition of OOTA is of course important, but those of us
writing code in the real world must rely on actual C++ implementations.
And in these implementations, this prohibition is in fact enforced by TSO
ordering in strongly ordered systems and by data dependency ordering in
weakly ordered systems.
Real-world implementations cannot store something until after they load
it, and this \emph{semantic dependency} results in compilers emitting
instructions such that the hardware enforcement applies.\footnote{
	The need to prohibit simple OOTA is one reason why compiler-based
	value speculation optimizations are frowned upon.}
Roughly speaking, there is semantic dependency from a given load to
a given store when a change in the value loaded that can change the
value stored.

\subsection{OOTA: rf vs. rfe}
\label{sec:OOTA: rf vs. rfe}

However, the ordering implied by the semantic dependency is only one
type of link in an OOTA cycle.
The other link extends from a given store to the load that returns the
value stored.
It is tempting to make the intuitive argument that this store must precede
this load in global time, and then to combine this with the notion that
any real C++ implementation must consume global time when computing a
semantic dependency.
This combination suggests that such implementations cannot exhibit
OOTA cycles.
This combination has been formalized by defining an OOTA cycle as a cycle
in sdep $\cup$ rf, where sdep is the set of semantic dependencies within
each thread and rf is the set of store-to-load links, whether intra-thread
(rfi) or between threads (rfe).\footnote{
	See Appendix~\ref{app:Inter-Process Communications} for definitions
	and properies of rf, rfe, and rfi.}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3-2-proc.litmus @@
\caption{OOTA With an Intra-Process Store-Load Link}
\label{lst:OOTA With an Intra-Process Store-Load Link}
\end{listing}

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/oota-3-2-proc-opt.litmus @@
\caption{OOTA With an Intra-Process Store-Load Link, Optimized}
\label{lst:OOTA With an Intra-Process Store-Load Link, Optimized}
\end{listing}

This is a fine definition, and is consistent with the words in the C++
standard, except for code fragments exemplified by
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link}.
This is an elaboration of
Listing~\ref{lst:Simple OOTA}
that adds \co{z} and adds lines~15 and~16 to \co{thread2()}.
The problem is that the C++ implementation is within its rights to
note that line~16 might well execute immediately after line~15, so that
no other thread had a chance to modify \co{z}.
Such an implementation might therefore behave as if the source code
had instead been as shown in
Listing~\ref{lst:OOTA With an Intra-Process Store-Load Link, Optimized},
with line~16 optimized away in favor of line~17 storing the value
in \co{r2} that was loaded from \co{y}.
And given that the load from \co{z} no longer exists, it cannot possibly
act as a temporal constraint.

In order to avoid these backwards-in-time rfi links, we substitute rfe
for rf, thus defining an OOTA cycle as a cycle in sdep $\cup$ rfe (or
\co{acyclic(sdep ; rfe)} in \co{herd7} notation), where rfe is the set
of store-to-load links, but excluding intra-thread rfi links, which are
instead considered to be part of sdep.
Although this does shunt additional complexity onto the term
``semantic dependency'', it also enables us to cleanly separate
the inter-thread and intra-thread portions of any given OOTA cycle.
% \footnote{
%	Thanks to Alan Stern for providing a litmus test that demonstrated
%	that we had unwittingly (but productively!) shifted our definition
%	to sdep $\cup$ rfe.}
@@@ Please see
Appendix~\ref{app:Informal Definition of Semantic Dependency}
for more a more detailed (albeit quite informal) definition.

@@@ Because there is a cycle, on a real system, at least one of the links
must be atemporal, that is, either a load returns a value before
that value is stored, or a store's address or value is computed before
a prior load returns a value used in that computation.

@@

@@@ The C++ standard says this in 33.5.4p8
(\co{[atomics.order]})~\cite{ThomasKoeppe2023N4950}:

@@

@@

\subsubsection{Simple Reordering as Litmus Test}
\label{app:Simple Reordering as Litmus Test}

Listing~\ref{lst:Simple Reordering as Litmus Test}
shows how
Listing~\ref{lst:Simple Reordering}
can be recast into a \co{herd7} litmus test.

\begin{listing}[tbp]
@@ DisplayRunLitmus litmus/simple-reordering.litmus @@
\caption{Simple Reordering as Litmus Test}
\label{lst:Simple Reordering as Litmus Test}
\end{listing}

@@

Although the examples presented in
Section~\ref{sec:OOTA: rf vs. rfe}
demonstrate that rf can be atemporal, they demonstrate neither
OOTA cycles nor OOTA-like behavior in full C++ implementations.
This appendix provides more involved examples that can be optimized
so as to demonstrate OOTA-like behavior.

Listing~\ref{lst:Why rfe Instead of Tried-And-True rf?}
revises
Listing~\ref{lst:Three-Process Version of JMM Causality Test Case 4 With Max and Min}
so that the value stored to \co{y} is forced to be at most the
value 17, and the value stored to \co{z} is forced to be at least the
value 17.
This of course means that the store to \co{z} must always be the value
17, as can be seen from the \co{herd7} output.
However, a single-thread-at-a-time analysis could not prove this.
